Instability and inequality in the COVID-19 lockdown crime drop

### 2021-03-04 draft_paper.Rmd
crime type profiles

### Crime data

@@

### Unit of analysis

@@

### Open Street Map

@@

## Method

@@

### Macro-level descriptives

@@

### Meso-level clustering

@@

We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions using three, four, five and six clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).

@@

### Cluster characteristics

@@

# Spatial distribution

@@

# Opportunity structure

@@

By way of example, there are sparse opportunities for crime in the largest and most stable cluster A. Most LSOA in this cluster have no nightlife facilities, no shops and no bicycle parking. Public transport is available but it is not common: the median LSOA in cluster A only contained four bus stops or railways stations. As we move along to medium and high crime clusters, which had higher pre-existing crime levels, and thus steeper declines during lockdown, these counts markedly increase. For instance, LSOA in cluster E contain a median of 21 nightlife facilities, 28 shops, 24 public transport nodes and 13 bicycle parking spaces. These counts increase further for cluster F, although we note that the cluster contains only nine LSOA.

@@

# Crime type profile

- For the purposes of simplicity and setting a benchmark, we have been using a measure of 'notifiable offences excluding drugs' for analysis.
- That said, we might expect the clusters identified to have different crime type profiles.
- Given the differing opportunity structures between clusters, the 'typical' crime profiles of each cluster may change considerably as lockdown commences.
- Alternatively, the pervasiveness of lockdown on mobility may have impacted on crime types uniformly within each cluster, in which case, we may see stability in crime type profiles before and after lockdown.
- Figure 5 summarises the profile of each cluster by calculating the proportion of total crime in each cluster.
- We observe considerable stability in the distribution of crime types both within and between clusters.
- Violence and sexual offences represent the most prevalent crime type across all clusters. The proportion of total crime in each cluster attributable to violence and sexual offences increases in the first two months of lockdown.
- This by no means implies that violence and sexual offences increased, but rather, that violence and sexual offences declined at a slower rate than other crime types, increasing its share of all crimes.
- In cluster F, we see a unique pattern, whereby theft from the person and other theft are squeezed considerably into April and May, while the share of public order increased.
- This might reflect the opportunity structures of LSOA in cluster F being weighted heavily towards theft (e.g. lots of public transport nodes).
- At the same time, the increase share in public order may reflect other activity, such as the protests around BLM and anti-lockdown movements, which take place in city centres for coverage.

>>>>

## Crime data

@@

## Unit of analysis

@@

## Open Street Map

@@

# Method

@@

## Macro-level descriptives

@@

## Meso-level clustering

@@

We deploy k-means using the `kml` package [@kmlpack] in R on notifiable offences (excluding drugs) counts for LSOA in the study region (N = 33,075) between February and August 2020. To achieve a parsimonious cluster solution while minimizing the risk of missing underlying variation, we proposed potential solutions between two and eight clusters respectively, choosing the final solution based on the Calinski-Criterion [@calinski1974dendrite]. For each potential solution, twenty redraws with different starting conditions were run to ensure that solutions were stable (see Appendix).

@@

## Cluster characteristics

@@

### Spatial distribution

@@

### Opportunity structure

@@

By way of example, there are sparse opportunities for crime in the largest and most stable cluster A. Most LSOA in this cluster have no nightlife facilities, no shops and no bicycle parking. Public transport is available but it is not common: the median LSOA in cluster A only contained four bus stops or railways stations. As we move along to medium and high crime clusters, which had higher pre-existing crime levels, and thus steeper declines during lockdown, these counts markedly increase. For instance, LSOA in cluster E contain a median of 21 nightlife facilities, 28 shops, 24 public transport nodes and 13 bicycle parking spaces. These counts increase further for cluster F, although we note that the cluster contains only nine LSOA. This is consistent with earlier findings which suggest that the crime decline most in these clusters: areas previously rich in opportunity, in which crime was pervasive, became the drivers of the lockdown crime drop.

@@

### Crime type profile

Given the unique longitudinal trends, spatial patterning and opportunity structure of the clusters identified using the aggregate notifiable offences measure, we might expect the clusters identified to have distinct crime type profiles. That is, some crime types may be more or less prevalent in certain clusters to begin with and / or there may be variation in the extent to which lockdown impeded opportunities across crime types. To explore this. Figure 6 summarises the crime type profile of each cluster.

In February, before the restrictions on mobility and social interaction, the crime type profiles of each cluster were already distinct. Higher crime clusters (e.g. E, and F) were weighted heavily towards shoplifting and theft. Low crime and stable clusters, by contrast, had higher proportions of criminal damage and arson, burglary and violence and sexual offences.

Overall, we observe a degree stability in the distribution of crime types both within and between clusters. The largest, stable clusters (A and B) have remarkably similar crime type profiles. Before lockdown, for each of these clusters, the most prevalent crimes were violence and sexual offences, both comprising 41% of total crime. This proportion increase on lockdown commencement, largely at the expense of burglary and vehicle crime, which decreased as a proportion of total crime. In clusters A and B, and indeed across all clusters, the proportion of total crime attributable to public order also increased.  These increases may reflect a seasonal effect: crimes such as violence and sexual offences and public order tend to increase between February and August in typical times, while vehicle crime and burglary are usually stable (@langton2021cs; see Appendix). This does not, by any means, indicate that these crimes increased during lockdown -- counts declined considerably -- but rather, these crimes declined less steeply relative to other offence categories. In the case of public order, this may be a result of lockdown-specific activity. Public order includes offences relating to processions and assemblies, and thus may capture gatherings which violated COVID-19 guidelines [@crown2021], and protests such as those relating to Black Lives Matter. Increases in public order offences across these areas would slow and work against a fall in a more typical public order crimes.

Notable disparities between clusters arise with regards to theft from the person and other theft in cluster F, which as reported earlier, drove a considerable part of the nationwide decline in crime. Theft in these LSOA was squeezed upon lockdown commencement. In February, theft crime types comprised 53% of total crime, but by April, this figure had dropped to 13%. This occurred despite these crime types typically remaining stable during equivalent times of year (@langton2021cs; see Appendix). Somewhat surprisingly, given the spatial distribution and opportunity structures of clusters E and F in city centres, shoplifting did not experience a similar squeeze in these LSOA.
